---
title: "Final Project Code"
author: "E. Morra, S. Miller"
date: "2023-11-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Libraries 
```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(tidymodels)
library(gghighlight)
library(knitr)
set.seed(1234)
```

San Francisco Data Set
Data Set has "San Francisco city employee salary data(gender column included)"
```{r}
#read in file
SF_dataset <- read.csv("SF_salary_data_gender.csv")
#data clean
SF_dataset <- SF_dataset[, !(colnames(SF_dataset) %in% c("Benefits", "TotalPayBenefits", "Status", "X"))]
SF_dataset <- SF_dataset[SF_dataset$gender != "", ]
SFmales <- sum(SF_dataset$gender == "m")
SFfemales <- sum(SF_dataset$gender == "f")
print(colnames(SF_dataset))
print(paste("The amount of males in the data set is is:", SFmales))
print(paste("The amount of females in the data set is is:", SFfemales))
```
Analyze and visualize cleaned San Francisco Data Set Summary
```{r}
options(scipen = 600000)
ggplot(SF_dataset, aes(x = gender, y = TotalPay, fill = gender)) +
  geom_boxplot() +
  labs(title = "Total Pay by Gender in San Francisco",
       x = "Gender",
       y = "Total Pay")
```

```{r}
mean_totalpay_f<- mean(SF_dataset$TotalPay[SF_dataset$gender == "f"], na.rm = TRUE)
mean_totalpay_f<- round(mean_totalpay_f, 2)

mean_totalpay_m<- mean(SF_dataset$TotalPay[SF_dataset$gender == "m"], na.rm = TRUE)
mean_totalpay_m<- round(mean_totalpay_m, 2)

print(paste("The mean total pay for females is:", mean_totalpay_f))
print(paste("The mean total pay for males is:", mean_totalpay_m))

max_totalpay_f<- max(SF_dataset$TotalPay[SF_dataset$gender == "f"], na.rm = TRUE)
max_totalpay_f<- round(max_totalpay_f, 2)

max_totalpay_m<- max(SF_dataset$TotalPay[SF_dataset$gender == "m"], na.rm = TRUE)
max_totalpay_m<- round(max_totalpay_m, 2)

print(paste("The maximum total pay for females is:", max_totalpay_f))
print(paste("The maximum total pay for males is:", max_totalpay_m))

```

```{r}
ggplot(SF_dataset, aes(x = Year, y = TotalPay, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Histogram of Total Pay by Gender and Year",
       x = "Total Pay",
       y = "year")
```


Texas Government Salary Data Set
```{r}
#read in file
texas_gov <- read.csv("texas_salaries.csv")
#data clean
remove_cols <- c("AGY", "MI", "CLASS.CODE", "STATE.NUMBER", "duplicated",
                 "multiple_full_time_jobs", "combined_multiple_jobs", "summed_annual_salary",
                 "hide_from_search")
texas_gov<- texas_gov[, !(names(texas_gov) %in% remove_cols)]

texas_gov <- texas_gov[complete.cases(texas_gov), ]

print(unique(texas_gov$GENDER))
texasmales <- sum(texas_gov$GENDER == "MALE           ")
texasfemales <- sum(texas_gov$GENDER == "FEMALE         ")
texas_gov <- na.omit(texas_gov)

print(colnames(texas_gov))
print(paste("The amount of males in the data set is is:", texasmales))
print(paste("The amount of females in the data set is is:", texasfemales))
```

```{r}
ggplot(texas_gov, aes(x = ETHNICITY, y = ANNUAL, fill = GENDER)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Annual Salaries by Gender and Ethnicity in the Texas Government",
       x = "Ethnicity",
       y = "Annual Salary")
```
Multiple Linear Regression
```{r}
print(unique(texas_gov$GENDER))
print(unique(texas_gov$ETHNICITY))
texas_gov$gender_num <- as.integer(factor(texas_gov$GENDER, levels = c("MALE           ", "FEMALE         ")))
texas_gov$ethnicity_num <- as.integer(factor(texas_gov$ETHNICITY, levels = c("WHITE          ", "BLACK          ", "HISPANIC       ", "OTHER          ", "AM INDIAN      ", "ASIAN          ")))

texas_annual <- lm(ANNUAL ~ gender_num + ethnicity_num, texas_gov)
names(texas_annual)
coefficients(texas_annual)
summary(texas_annual)
```

Testing / training - see lab 10

- Create an initial split

```{r}
tex_tt <- texas_gov[, -(1:4)]
tex_tt <- tex_tt[, -(4:7)]
tex_tt <- tex_tt[, -(5:6)]
# take away unneccesary vars
set.seed(1122)
texas_split <- initial_split(tex_tt) # prop = 3/4 by default
```

- Save training data
```{r}
texas_train <- training(texas_split)
dim(texas_train)
```

- Save testing data
```{r}
texas_test  <- testing(texas_split)
dim(texas_test)
```
##### Specify model

```{r}
texas_mod <- linear_reg() %>%
  set_engine("lm")

texas_mod
```
##### Build recipe


```{r}
texas_rec <- recipe(ANNUAL ~ ., data = texas_train) %>%
  # remove zero variance predictors
  step_zv(all_predictors())
```

```{r}
texas_rec
```
##### Build workflow

```{r}
texas_wflow <- workflow() %>%
  add_model(texas_mod) %>%
  add_recipe(texas_rec)
```

```{r}
texas_wflow
```
##### Fit model


```{r}
texas_fit <- texas_wflow %>%
  fit(data = texas_train)
```


```{r}
tidy(texas_fit) %>%
  print(n = 12)
```

#### Evaluate model



##### Make predictions for training data

```{r}
texas_train_pred <- predict(texas_fit, texas_train) %>%
  bind_cols(texas_train %>% select(ANNUAL, GENDER))

texas_train_pred
```
##### R-squared

Percentage of variability in salary explained by the model

```{r}
rsq(texas_train_pred, truth = ANNUAL, estimate = .pred)
```



```{r}
rmse(texas_train_pred, truth = ANNUAL, estimate = .pred)
```


Are models with high or low RMSE are more preferable?

##### Interpreting RMSE


Is this RMSE considered low or high?

```{r}
texas_train %>%
  summarise(min = min(ANNUAL), max = max(ANNUAL)) 
```



##### Make predictions for testing data

```{r}
texas_test_pred <- predict(texas_fit, texas_test) %>%
  bind_cols(texas_test %>% select(ANNUAL, GENDER))

texas_test_pred
```



##### Evaluate performance on testing data

- RMSE of model fit to testing data

```{r}
rmse(texas_test_pred, truth = ANNUAL, estimate = .pred)
```

- $R^2$ of model fit to testing data

```{r}
rsq(texas_test_pred, truth = ANNUAL, estimate = .pred)
```



##### Training vs. testing



```{r echo=FALSE}
rmse_train_texas <- rmse(texas_train_pred, truth = ANNUAL, estimate = .pred) %>% 
  pull(.estimate) %>%
  round(3)

rsq_train_texas <- rsq(texas_train_pred, truth = ANNUAL, estimate = .pred) %>% 
  pull(.estimate) %>%
  round(3)

rmse_test_texas <- rmse(texas_test_pred, truth = ANNUAL, estimate = .pred) %>% 
  pull(.estimate) %>%
  round(3)

rsq_test_texas <- rsq(texas_test_pred, truth = ANNUAL, estimate = .pred) %>% 
  pull(.estimate) %>%
  round(3)

tibble(
  data      = c(rep("train", 2), rep("test", 2)),
  estimate  = c(rmse_train_texas, rsq_train_texas, rmse_test_texas, rsq_test_texas),
  metric    = c("RMSE", "R-squared", "RMSE", "R-squared")
) %>%
  pivot_wider(names_from = data, values_from = estimate) %>%
  bind_cols(comparison = c("RMSE lower for training", "R-squared higher for training")) %>%
  kable()
```







IT Specialists Data Set
## Read in data set 
```{r}
it_specialists <- read.csv("Salaries-of-IT-specialists.csv")

names(it_specialists)[5] <- "Years of Experience"
names(it_specialists)[6] <- "Annual Salary"
names(it_specialists)[10] <- "Employment Type"
names(it_specialists)[11] <- "Number of Projects"
names(it_specialists)[12] <- "Job Satisfaction"

it <- subset(it_specialists, select = -X)
its <- subset(it, select = -ID)

its <- its[1:218, ]

print(unique(its$Gender))
itsmales <- sum(its$Gender == "Male")
itsfemales <- sum(its$Gender == "Female")
print(paste("The amount of males in the data set is is:", itsmales))
print(paste("The amount of females in the data set is is:", itsfemales))
```

## Histogram for male and female within each continent
```{r}
ggplot(its, aes(x = Continent, y = `Annual Salary`, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Annual Salary of Genders For Each Continent",
       x = "Continent",
       y = "Salary",
       fill = "Gender") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_discrete(name = "Gender")
```


## Multiple linear regression using columns as predictors for salary 
```{r}
its$Education_numeric <- as.integer(factor(its$Education, levels = c("None", "Bachelor's", "Master's")))
its$Gender_numeric <- as.integer(factor(its$Gender, levels = c("Male", "Female")))

salaries <- lm(`Annual Salary` ~ `Years of Experience`+ Gender_numeric + Age + Education_numeric, its)
names(salaries)
coefficients(salaries)
summary(salaries)
```
Testing / training - see lab 10

- Create an initial split

```{r}
set.seed(1122)
its_split <- initial_split(its) # prop = 3/4 by default
```

- Save training data
```{r}
its_train <- training(its_split)
dim(its_train)
```

- Save testing data
```{r}
its_test  <- testing(its_split)
dim(its_test)
```
##### Specify model

```{r}
its_mod <- linear_reg() %>%
  set_engine("lm")

its_mod
```
##### Build recipe


```{r}
its_rec <- recipe(`Annual Salary` ~ ., data = its_train) %>%
  # remove zero variance predictors
  step_zv(all_predictors())
```

```{r}
its_rec
```
##### Build workflow

```{r}
its_wflow <- workflow() %>%
  add_model(its_mod) %>%
  add_recipe(its_rec)
```

```{r}
its_wflow
```
##### Fit model


```{r}
its_fit <- its_wflow %>%
  fit(data = its_train)
```


```{r}
tidy(its_fit) %>%
  print(n = 12)
```

#### Evaluate model



##### Make predictions for training data

```{r}
its_train_pred <- predict(its_fit, its_train) %>%
  bind_cols(its_train %>% select(`Annual Salary`, Gender))

its_train_pred
```
##### R-squared

Percentage of variability in salary explained by the model

```{r}
rsq(its_train_pred, truth = `Annual Salary`, estimate = .pred)
```



```{r}
rmse(its_train_pred, truth = `Annual Salary`, estimate = .pred)
```


Are models with high or low RMSE are more preferable?

##### Interpreting RMSE


Is this RMSE considered low or high?


```{r}
its_train %>%
  summarise(min = min(`Annual Salary`), max = max(`Annual Salary`)) 
```



##### Make predictions for testing data

```{r}
its_test_pred <- predict(its_fit, its_test) %>%
  bind_cols(its_test %>% select(`Annual Salary`, Gender))

its_test_pred
```



##### Evaluate performance on testing data

- RMSE of model fit to testing data

```{r}
rmse(its_test_pred, truth =`Annual Salary`, estimate = .pred)
```

- $R^2$ of model fit to testing data

```{r}
rsq(its_test_pred, truth = `Annual Salary`, estimate = .pred)
```



##### Training vs. testing



```{r echo=FALSE}
rmse_train_its <- rmse(its_train_pred, truth = `Annual Salary`, estimate = .pred) %>% 
  pull(.estimate) %>%
  round(3)

rsq_train_its <- rsq(its_train_pred, truth = `Annual Salary`, estimate = .pred) %>% 
  pull(.estimate) %>%
  round(3)

rmse_test_its <- rmse(its_test_pred, truth = `Annual Salary`, estimate = .pred) %>% 
  pull(.estimate) %>%
  round(3)

rsq_test_its <- rsq(its_test_pred, truth = `Annual Salary`, estimate = .pred) %>% 
  pull(.estimate) %>%
  round(3)

tibble(
  data      = c(rep("train", 2), rep("test", 2)),
  estimate  = c(rmse_train_its, rsq_train_its, rmse_test_its, rsq_test_its),
  metric    = c("RMSE", "R-squared", "RMSE", "R-squared")
) %>%
  pivot_wider(names_from = data, values_from = estimate) %>%
  bind_cols(comparison = c("RMSE lower for training", "R-squared higher for training")) %>%
  kable()
```
 
 
 ## ^ taken from lab 10** 
 
 
 White House Data Set
 
```{r}
white_house <- read.csv("wh_staff_dataset.csv")
glimpse(white_house)
```
 
 
 
```{r}
# remove NAs to check for max salary for box plot axis 
colSums(is.na(white_house))
wh <- na.omit(white_house)
colSums(is.na(wh))

whhs <- wh[wh$salary != 0, ]

max(whhs$salary)

# remove data with blank gender entry for boxplot 
print(unique(whhs$gender))

wh2 <- whhs[whhs$gender != "", ]

print(unique(wh2$gender))
whmales <- sum(wh2$gender == "Male")
whfemales <- sum(wh2$gender == "Female")
print(paste("The amount of males in the data set is is:", whmales))
print(paste("The amount of females in the data set is is:", whfemales))

options(scipen = 250000)
ggplot(wh2, aes(x = gender, y = salary, fill = gender)) +
  geom_boxplot() +
  labs(title = "Salary Depending on Gender",
       y = "Salary",
       x = "Gender") 
```



```{r}
mean_salaries <- aggregate(salary ~ gender, data = wh2, FUN = mean)

mean_salaries

median_salaries <- aggregate(salary ~ gender, data = wh2, FUN = median)

median_salaries

```



# line graph for male and female within each year
```{r}
ggplot(wh2, aes(x = year, y = salary, color = gender)) +
  geom_line() +
  labs(title = "Annual Salary Each Year for Males and Females",
       x = "Year",
       y = "Annual Salary",
       color = "Gender") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_discrete(name = "Gender")
```



```{r}
# max salary male or female 
max_row <- wh2[which.max(wh2$salary), ]
max_row

```

```{r}
print(unique(wh$status))
print(unique(wh$pay_basis))

# changing categorical vars 

wh2$gender_num <- as.integer(factor(wh2$gender, levels = c("Male", "Female")))
wh2$status_num <- as.integer(factor(wh2$status, levels = c("Employee", "Detailee", "Employee (part-time)")))
wh2$pay_num <- as.integer(factor(wh2$pay_basis, levels = c("Per Annum", "Per Diem")))

# multiple linear regression 

salaries2 <- lm(salary ~ gender_num + year + pay_num + status_num, wh2)
names(salaries2)
coefficients(salaries2)
summary(salaries2)

```



Testing / training - see lab 10

- Create an initial split

```{r}
wh_tt <- wh2[, -(1:2)]
wh_tt <- wh_tt[, -(5:8)]
# take away names, positions (too many to count), year, and num cat vars
set.seed(1122)
wh_split <- initial_split(wh_tt) # prop = 3/4 by default
```

- Save training data
```{r}
wh_train <- training(wh_split)
dim(wh_train)
```

- Save testing data
```{r}
wh_test  <- testing(wh_split)
dim(wh_test)
```
##### Specify model

```{r}
wh_mod <- linear_reg() %>%
  set_engine("lm")

wh_mod
```
##### Build recipe


```{r}
wh_rec <- recipe(salary ~ ., data = wh_train) %>%
  # remove zero variance predictors
  step_zv(all_predictors())
```

```{r}
wh_rec
```
##### Build workflow

```{r}
wh_wflow <- workflow() %>%
  add_model(wh_mod) %>%
  add_recipe(wh_rec)
```

```{r}
wh_wflow
```
##### Fit model


```{r}
wh_fit <- wh_wflow %>%
  fit(data = wh_train)
```


```{r}
tidy(wh_fit) %>%
  print(n = 12)
```

#### Evaluate model



##### Make predictions for training data

```{r}
wh_train_pred <- predict(wh_fit, wh_train) %>%
  bind_cols(wh_train %>% select(salary, gender))

wh_train_pred
```
##### R-squared

Percentage of variability in salary explained by the model

```{r}
rsq(wh_train_pred, truth = salary, estimate = .pred)
```



```{r rmse-train}
rmse(wh_train_pred, truth = salary, estimate = .pred)
```


Are models with high or low RMSE are more preferable?

##### Interpreting RMSE


Is this RMSE considered low or high?


```{r ref.label="rmse-train"}
```



```{r}
wh_train %>%
  summarise(min = min(salary), max = max(salary)) 
```



##### Make predictions for testing data

```{r}
wh_test_pred <- predict(wh_fit, wh_test) %>%
  bind_cols(wh_test %>% select(salary, gender))

wh_test_pred
```



##### Evaluate performance on testing data

- RMSE of model fit to testing data

```{r}
rmse(wh_test_pred, truth = salary, estimate = .pred)
```

- $R^2$ of model fit to testing data

```{r}
rsq(wh_test_pred, truth = salary, estimate = .pred)
```



##### Training vs. testing



```{r echo=FALSE}
rmse_train_wh <- rmse(wh_train_pred, truth = salary, estimate = .pred) %>% 
  pull(.estimate) %>%
  round(3)

rsq_train_wh <- rsq(wh_train_pred, truth = salary, estimate = .pred) %>% 
  pull(.estimate) %>%
  round(3)

rmse_test_wh <- rmse(wh_test_pred, truth = salary, estimate = .pred) %>% 
  pull(.estimate) %>%
  round(3)

rsq_test_wh <- rsq(wh_test_pred, truth = salary, estimate = .pred) %>% 
  pull(.estimate) %>%
  round(3)

tibble(
  data      = c(rep("train", 2), rep("test", 2)),
  estimate  = c(rmse_train_wh, rsq_train_wh, rmse_test_wh, rsq_test_wh),
  metric    = c("RMSE", "R-squared", "RMSE", "R-squared")
) %>%
  pivot_wider(names_from = data, values_from = estimate) %>%
  bind_cols(comparison = c("RMSE lower for training", "R-squared higher for training")) %>%
  kable()
```